### 2020-06-04 dataclean.tex
combine view figs

% !TEX root = main.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Viewing and Cleaning the Data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[bt]
  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf}
  \caption{Serial number view of GPU life and failures. Legend: black
    dots are installs, black lines are lifetimes at installed
    location, blue squares are OTB events, red triangles are DBE
    events, and black ] are ``last seen'' events.}
  \label{fig:gpuview}
\end{figure*}
\begin{figure*}[tb]
  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf}
  \caption{Location view of GPU life and failures.  Legend: black dots
    are installs, black lines are lifetimes at installed location,
    blue squares are OTB events, red triangles are DBE events, and
    black ] are ``last seen'' events.}
  \label{fig:locview}
\end{figure*}
\section{Viewing and Cleaning the Data}
\label{section:dataclean}
As we need to recover durations of GPU operation from this data, correct
processing involves time adjustments for switching between daylight
saving time and standard time and leap years. We perform this by
setting a reference time zone (Eastern time) and converting all
date-times from strings into POSIX date-time variables with the R
\pkg{lubridate} package \cite{lubridate}, which enables appropriate
date arithmetic and sensible date constructs for graphs.

@@

To get some intuition for the GPU lifetimes on Titan, we give two
views of 60 randomly selected GPUs (\rev{Fig.}~\ref{fig:gpuview}) and 80
randomly selected locations (\rev{Fig.}~\ref{fig:locview}).
The GPU view
visually documents the life of each GPU unit: when it was installed and
removed at various locations, its DBE and OTB events, and the last
time it was seen. The location view documents the life of a location:
when different GPUs were installed and removed, their OTB and DBE
events, and whether a removal was the last time the unit was seen on
the system. These views were critical to understanding the data and to
verifying various data processing decisions.

The third rework cycle was used to label GPUs as {\em old} batch and
{\em new} batch. The two batches are clearly identifiable in the GPU
view of Fig.~\ref{fig:gpuview} as starting near 2017 timeframe. More
frequent OTB and DBE events are apparent in the old batch. It is also
clear from this view that practically all new GPUs stayed at their
initial install location whereas the old GPUs were occasionally
reinstalled at new locations. Nevertheless, a separate analysis
determined that the vast majority of time of the vast majority of
units is spent at one location by both the new and the old units.

>>>>

% !TEX root = main.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Viewing and Cleaning the Data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Viewing and Cleaning the Data}
\label{section:dataclean}
As we need to recover durations of GPU operation from this data, correct
processing involves time adjustments for switching between daylight
saving time and standard time and leap years. We perform this by
setting a reference time zone (Eastern time) and converting all
date-times from strings into POSIX date-time variables with the R
\pkg{lubridate} package \cite{lubridate}, which enables appropriate
date arithmetic and sensible date constructs for graphs.

@@

To get some intuition for the GPU lifetimes on Titan, we give two
views of \rev{80} randomly selected GPUs and \rev{80} randomly
selected locations (\rev{Fig.}~\ref{fig:gpu-view}).
\begin{figure*}[p]
  \begin{center}
  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_sn.pdf} \\
  (a) \\[1em]
  \includegraphics[trim=0 7 0 5,clip,width=6.5in]{figs/sample_loc.pdf} \\
  (b)
  \end{center}
  \caption{\rev{Serial number view (a) and location view (b) of GPU
      life and failures. Both SN and locations are randomly
      selected. Legend: black dots are installs, black lines are
      lifetimes at installed location, blue squares are OTB events,
      red triangles are DBE events, and black ] are ``last seen''
      events.}}
  \label{fig:gpu-view}
\end{figure*}
The GPU view
visually documents the life of each GPU unit: when it was installed and
removed at various locations, its DBE and OTB events, and the last
time it was seen. The location view documents the life of a location:
when different GPUs were installed and removed, their OTB and DBE
events, and whether a removal was the last time the unit was seen on
the system. These views were critical to understanding the data and to
verifying various data processing decisions.

The third rework cycle was used to label GPUs as {\em old} batch and
{\em new} batch. The two batches are clearly identifiable in the
\rev{SN} view of Fig.~\ref{fig:gpu-view} as those starting near 2017
timeframe. More frequent OTB and DBE events are apparent in the old
batch. It is also clear from this view that practically all new GPUs
stayed at their initial install location whereas the old GPUs were
occasionally reinstalled at new locations. Nevertheless, a separate
analysis determined that the vast majority of time of the vast
majority of units is spent at one location by both the new and the old
units.

### 2020-06-04 intro.tex
updated Rmd and added silver corrosion ref
combine view figs

\begin{figure*}[tb]
  \centering
  \includegraphics[width=\textwidth]{figs/chronology001.pdf}
  \caption{\rev{Volume of GPU swaps detected on Titan at individual
    inventories (narrow blue) and yealy sum totals for 2014 and later
    (wide gray) over its entire lifetime. High swap volumes prior to
    2014 reflect two major rework cycles on mechanical assembly issues
    at PCIe connectors that can be characterized as a break-in
    period. Our analysis in this paper excludes the break-in period
    and is focused on units in place by the end of 2013 or swapped in
    during the years that follow}.}
  \label{fig:chronology}
\end{figure*}
Figure~\ref{fig:chronology} illustrates the chronology of the rework
cycles and stable periods as indicated by the number of GPU changes at
periodic inventories. The first two rework cycles before 2014
involving blade mechanical assemblies can be characterized as a
break-in period on a new system that is first of its kind and pushes
many technological boundaries. \rev{This early period included
  extensive field engineering and experimentation to determine cause,
  down times, and large temporal and completeness gaps in inventory
  runs. We include the early swap data in this figure only for
  completeness and to underscore the amount of work required to field
  a world's largest supercomputer. However, we exclude this period
  from further analysis because we have a gap in explanation for one
  of the swap spikes and because the separation is made possible by
  the second, late 2013, GPU rework cycle.} Late 2013 begins a very
long period of high reliability and stable operation with very few
issues until about mid 2016, when GPU failures begin to rise. This
results in the final rework cycle, replacing 11,000 GPUs throughout
much of 2017.

Finding the root cause for the 2016 failures took a great deal of
testing and involvement of materials science and microscopy
researchers. The failures were traced to a resistor on the GPU
circuit board (not the GPU chip iself). Throughout this paper, when we
refer to replacing the GPU, we refer to the entire circuitboard along
with its GPU chip.

>>>>

\begin{figure*}[tb]
  \centering
  \includegraphics[width=\textwidth]{figs/chronology001.pdf}
  \caption{\rev{Volume of GPU swaps detected on Titan at individual
    inventories (narrow blue) and yealy sum totals for 2014 and later
    (wide gray) over its entire lifetime. High swap volumes prior to
    2014 reflect two major rework cycles on mechanical assembly issues
    at PCIe connectors that can be characterized as a break-in
    period. Our analysis in this paper excludes the break-in period
    and is focused on units in place by the end of 2013 or swapped in
    during the years that follow}.}
  \label{fig:chronology}
\end{figure*}
Figure~\ref{fig:chronology} illustrates the chronology of the rework
cycles and stable periods as indicated by the number of GPU changes at
periodic inventories. The first two rework cycles before 2014
involving blade mechanical assemblies can be characterized as a
break-in period on a new system that is first of its kind and pushes
many technological boundaries. \rev{This early period included
  extensive field engineering and experimentation, down times, as well
  as temporal and completeness gaps in inventory runs due to down
  times. We include the early swap data in this figure only for
  completeness and to underscore the massive amount of hardware work
  required to field a world's largest supercomputer. However, as we
  focus on GPU operation after the second rework cycle we exclude this
  early period from further analysis.} Late 2013 begins a very long
period of high reliability and stable operation with very few issues
until about mid 2016, when GPU failures begin to rise. This results in
the final rework cycle, replacing 11,000 GPUs throughout much of 2017.

Finding the root cause for the 2016 failures took a great deal of
testing and the involvement of materials science and microscopy
researchers. The failures were traced to a resistor on the GPU circuit
board (not the GPU chip iself) \fix{Was anything published about this?
  If not, let's use \cite{Vargas2009TheCO}}. Throughout this paper,
  when we refer to replacing the GPU, we refer to the entire
  circuitboard along with its GPU chip.

### 2020-06-04 tbf.tex
combine view figs

Apart from the center cluster, many GPUs also have a very 
low MTBF. This failure characteristic is mainly an artifact from
troubleshooting a period of increasing failures in mid to late 2016.
After a GPU experienced an OTB or a DBE a second reboot was attempted at
the same or possibly different location. Usually, but not always, a
second OTB or DBE was generated immediately or possibly after a short
time. When moved to a new location, this generates an artifact in our
data with an unusually short MTBF because we use the {\em insert} time as
a reference for the first failure on a relocated GPU. This drawback is
balanced by the advantage that this way we can discount out of service times
which were frequent particularly during early 2017 as can be seen in
Fig.~\ref{fig:locview}.

>>>>

Apart from the center cluster, many GPUs also have a very 
low MTBF. This failure characteristic is mainly an artifact from
troubleshooting a period of increasing failures in mid to late 2016.
After a GPU experienced an OTB or a DBE a second reboot was attempted at
the same or possibly different location. Usually, but not always, a
second OTB or DBE was generated immediately or possibly after a short
time. When moved to a new location, this generates an artifact in our
data with an unusually short MTBF because we use the {\em insert} time as
a reference for the first failure on a relocated GPU. This drawback is
balanced by the advantage that this way we can discount out of service times
which were frequent particularly during early 2017 as can be seen in
the location view of Fig.~\ref{fig:gpu-view}.
